\propertitle

### 2019-07-25 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-22T13:43:19+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Answering this kind of questions always engages an element of uncertainty.
Our answers therefore involve experimental data, such as neuronal
recordings from specific brain areas, and probabilities or \dobs, based on
prior knowledge, about biological conditions and mechanisms that cannot be
experimentally ascertained. Such \dobs\ are often formulated as simplified
\enquote{models} to be mathematically more tractable.

Despite remarkable advances in recording technologies, the best
experimental measurements of neuronal activity can still only record a very
small sample of neurons compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only, somehow neglecting, in their assumptions, that the
recorded neurons are a sample from a larger network. This kind of isolation
assumptions sometimes escape attention, being subtly hidden in the
mathematics. Some probabilistic models try to take also unrecorded neurons
into account, but become very complex in doing so.

In the present work we give an answer to this question: How much can the
total activity of a large neuronal network be, if we have observed the
activity of a very small sample thereof? We'll quantify our \dobs\ about
the possible answers by combining, in a straightforward way, the
maximum-entropy method and basic sampling relations of the probability
calculus. Later on we'll show that our \dobs\ can be quantified by
exclusively using the probability calculus. This derivation will provide a
more accurate quantification, revealing that the maximum-entropy answer is
only a first approximation.

We apply our approach to two concrete data sets: the activity of 65 neurons
recorded from a rat's Medial Entorhinal Cortex \mynote{ref}, and the
activity of 159 neurons recorded from a macaque's Motor Cortex
\mynote{ref}\mynote{add recording length}. In the first data set, the most
interesting finding is that the most probable frequency distribution for
the total activity of the full network has two very distinct modes, both at
low activities, %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
see \fig~\ref{fig:plots_6mom}. The analogous frequency distribution for the
second data set doesn't have two modes but still presents one prominent
shoulder in its one low-activity mode. Note that these guessed features of
the full network aren't observed in the sample, nor can they be inferred by
the application of maximum-entropy \emph{to the sample alone}.
\mynote{Monte Carlo sampling: say something about the deviation from the
  most probable frequency distribution}

@@

The maximum-entropy or minimum-relative-entropy method
\citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
This method is often used to test whether some statistics of the data, for
example second-order time correlations, is sufficient for quantifying our
\dob\ about some quantities of the system. It can be considered as an
approximation of probabilistic models based on various assumptions of
inferential sufficiency \citep{jaynes1986d_r1996,portamana2017}.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-25T11:36:58+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Answering this kind of questions always engages an element of uncertainty.
We can't say \enquote{the answer is such and such}, but can at best assign
a degree of reasonable belief -- that is, probability -- to every possible
answer. The assessment of this distribution of belief involves experimental
data, such as recordings of neuronal activity from specific brain areas,
and pre-data knowledge about biological conditions and mechanisms. Our
pre-data \dobs\ are often simplified and called \enquote{models} to be
mathematically more tractable.

Despite remarkable advances in recording technologies, the best
experimental measurements of neuronal activity can still only record a very
small sample of neurons compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only, somehow neglecting, in their assumptions, that the
recorded neurons are a sample from a larger network. This kind of isolation
assumptions sometimes escape attention, being subtly hidden in the
mathematics. Some probabilistic models try to take also unrecorded neurons
into account individually, and therefore become very complex\mynote{refs}.
It would be useful to have models that operate in between: addressing the
larger brain area where a sample comes from, but without adding too much
complexity and detail about it. Such intermediate models would be useful,
for example, for preliminary investigations, to help us to decide which
hypotheses to discard and which to consider for more complex and costly
studies.

In the present work we propose such an intermediate model. It answers this
question: How much can the \emph{total} activity of a large neuronal
network be, if we have observed the activity of a very small sample
thereof? This model addresses a larger brain area, avoiding the assumption
of isolation of the sample. And by focusing on the total activity, rather
than the activity of individual neurons, it remains simple and numerically
tractable.

The proposed model is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus. The maximum-entropy or minimum-relative-entropy method
\citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
This method has often been used to test whether a smaller part of the data,
for example time averages and second-order correlations, is sufficient for
quantifying our \dob\ about some aspects of the neuronal activity. We'll
consider this particular application also for our proposed model.
Maximum-entropy models can be considered as approximations of Bayesian
models based on various assumptions of inferential sufficiency
\citep{jaynes1986d_r1996,portamana2017}.
% Later on we'll show that our \dobs\ can be quantified by
% exclusively using the probability calculus. This derivation will provide a
% more accurate quantification, revealing that the maximum-entropy answer is
% only a first approximation.

We apply our proposed model to two concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded from a rat's Medial Entorhinal
  Cortex \citep{stensolaetal2012}, \item the activity of 159 neurons
  recorded from a macaque's Motor Cortex \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model yields***

In the first data set, the most
interesting finding is that the most probable frequency distribution for
the total activity of the full network has two very distinct modes, both at
low activities, %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
see \fig~\ref{fig:plots_6mom}. The analogous frequency distribution for the
second data set doesn't have two modes but still presents one prominent
shoulder in its one low-activity mode. Note that these guessed features of
the full network aren't observed in the sample, nor can they be inferred by
the application of maximum-entropy \emph{to the sample alone}.
\mynote{Monte Carlo sampling: say something about the deviation from the
  most probable frequency distribution}

@@
